---
title: "Coaching Networks and Career Pathways Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(igraph)
library(ggraph)
library(tidygraph)
library(lubridate)
```

# Overview

This analysis explores coaching networks and career pathways in NCAA women's basketball, focusing on which schools produce future head coaches and identifying coaching "trees."

Your Task: Work through each section, making decisions and identifying potential story angles from the data.

---

# Data Loading

Load the CSV file and check out the data to see what you're working with.

```{r load-data}
coaching <- read_csv("https://raw.githubusercontent.com/dwillis/wbb/refs/heads/main/ncaa/coaches/coaching_histories_merged.csv")

cat("\nDataset summary:\n")
cat("Total records:", nrow(coaching), "\n")
cat("Unique coaches:", n_distinct(coaching$coach), "\n")
cat("Unique schools:", n_distinct(coaching$college), "\n")
cat("Division I records:", sum(coaching$division == "I", na.rm = TRUE), "\n")
```

---

# Analysis 1: Head Coach Production Pipeline

## DECISION POINT 1: Defining "Future Head Coaches"

**Question for you**: Should we focus only on coaches who became D1 head coaches, or also include those who became head coaches at D2, D3, or other levels? How does this choice affect the story? Does it matter if someone became a D3 head coach vs. stayed in D1 as an assistant?

**I think choose all levels, so option two.** (modify the code below):

```{r head-coach-pipeline}
d1_head_coaches <- coaching |>
  filter(position_title_standardized == "Head Coach") |> 
  distinct(coach) |>
  pull(coach)

cat("Number of head coaches identified:", length(d1_head_coaches), "\n")
```


## Finding Previous Positions

```{r previous-positions}
# Find their previous positions (before becoming head coach)
future_hc_positions <- coaching |>
  filter(coach %in% d1_head_coaches, position_title_standardized != 'Head Coach') |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    is_head_coach = position_title_standardized == "Head Coach",
    first_hc_year = min(start_year[is_head_coach], na.rm = TRUE),
    before_hc = start_year < first_hc_year | is.na(first_hc_year)
  ) |>
  ungroup() 

cat("Total positions held by future head coaches:", nrow(future_hc_positions), "\n")
```

## DECISION POINT 2: Which Division Matters for Training?

**Question for you**: When looking at which schools produce future head coaches, should we only count their D1 experience, or also their D2/D3/other experience?

All of their experience. It all counts as coaching. 

```{r production-schools}

hc_production_schools <- future_hc_positions |> 
  filter(before_hc) |> 
  group_by(college_clean, position_title_standardized) |> 
  summarize(
    future_hcs = n_distinct(coach),
    .groups = "drop"
  ) |> 
  arrange(desc(future_hcs))
cat("\nTop 20 D1 schools producing future head coaches (all positions):\n")
hc_production_schools |>
  group_by(college_clean) |>
  summarize(total_future_hcs = sum(future_hcs), .groups = "drop") |>
  arrange(desc(total_future_hcs)) |>
  head(20) |>
  print(n = 20)
```

**Think about**: Does a stint at a D3 school early in someone's career "count" as part of their development?

100% it does

## Position Type Breakdown

```{r position-breakdown}
cat("\n\nTop schools by position type:\n")
hc_production_schools |>
  filter(position_title_standardized %in% c("Assistant Coach", "Associate Head Coach")) |>
  group_by(position_title_standardized) |>
  slice_max(future_hcs, n = 10) |>
  arrange(position_title_standardized, desc(future_hcs))
```

### 💡 STORY IDEAS - Section 1

**Based on what you see above, brainstorm 2-3 story angles:**

1. A handful of schools like Auburn, Mississippi State and Miami quietly operate as the NCAA’s biggest coaching “factories,” producing more future head coaches than many of the sport’s traditional power programs.
2. The data shows that the real coaching pipelines aren’t always bluebloods—programs like East Carolina, Cincinnati and Wake Forest have become unexpected launchpads for future head coaches.
3. The data makes one thing obvious: the clearest path to becoming a head coach still runs through years spent as an assistant, where most future head coaches actually build their resumes.

**Questions to consider:**
- Which school dominates the list? Is this surprising?
- Are there mid-major schools punching above their weight?
- Are there regional patterns?

---

# Analysis 2: Career Ladder Analysis

## DECISION POINT 3: Minimum Position Threshold

**Question for you**: Should we require a minimum number of positions before someone counts as having a "career ladder"? (e.g., exclude people who went straight from 1 assistant job to head coach?)

```{r career-ladder}
# STUDENT DECISION: Set minimum number of positions
MIN_POSITIONS <- 2

# Sequence of positions before becoming head coach
career_sequences <- future_hc_positions |>
  filter(before_hc) |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  summarize(
    positions_held = list(position_title_standardized),
    n_positions = n(),
    final_position = last(position_title_standardized),
    .groups = "drop"
  ) |>
  filter(n_positions >= MIN_POSITIONS)  # APPLIES YOUR THRESHOLD

cat("Coaches meeting threshold:", nrow(career_sequences), "\n")

cat("\nMost common final position before becoming D1 head coach:\n")
career_sequences |>
  count(final_position, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  head(10)

# Average number of positions held before head coach role
cat("\n\nAverage D1 positions held before becoming head coach:",
    round(mean(career_sequences$n_positions, na.rm = TRUE), 1))
```

## Common Career Pathways

```{r pathways}
# STUDENT DECISION: How many positions should we examine?
PATHWAY_LENGTH <- 3

cat("\n\nCommon", PATHWAY_LENGTH, "-position pathways to head coach:\n")
career_sequences |>
  filter(n_positions == PATHWAY_LENGTH) |>
  mutate(pathway = map_chr(positions_held, ~paste(.x, collapse = " → "))) |>
  count(pathway, sort = TRUE) |>
  head(15) |>
  print(n = 15)
```

### 💡 STORY IDEAS - Section 2

**Based on the career ladder data, brainstorm 2-3 story angles:**

1. The clearest pattern in the data is how often future head coaches stack multiple assistant stops, it’s basically the sport’s default ladder.
2. A surprising chunk of coaches climb through the exact same sequence, assistant to assistant to assistant, making the grind more of a requirement than an outlier.
3. Even the coaches who do “move up” aren't jumping rungs; most bounce between staff and assistant roles before ever getting a shot at running their own program.

**Questions to consider:**
- Are there "fast track" coaches who skipped steps?
- Does the Associate Head Coach position matter, or do people jump from Assistant to Head?
- How important is the "Staff" level as a starting point?

---

# Analysis 3: Coaching Trees and Mentorship Networks

## DECISION POINT 4: Time Overlap Requirements

**Question for you**: How should we define "worked under"? Should we require they overlapped for at least one full year, or is any overlap sufficient?

```{r coaching-trees}

MIN_OVERLAP_YEARS <- 0  

coaching_relationships <- coaching |>
  select(
    coach,
    college_clean,
    position_title_standardized,
    start_year,
    end_year
  ) |>
  

  left_join(
    coaching |>
      filter(position_title_standardized == "Head Coach") |>
      select(
        hc_name = coach,
        college_clean,
        hc_start = start_year,
        hc_end = end_year
      ),
    by = "college_clean",
    relationship = "many-to-many"
  ) |>
  

  filter(
    !is.na(hc_start),                   
    coach != hc_name,                     
    position_title_standardized %in% c(
      "Assistant Coach",
      "Associate Head Coach"
    )
  ) |>
  
  mutate(
    adj_end = coalesce(end_year, 2025),
    adj_hc_end = coalesce(hc_end, 2025),
    overlap_years = pmin(adj_end, adj_hc_end) - pmax(start_year, hc_start)
  ) |>
  
  filter(overlap_years >= MIN_OVERLAP_YEARS) |>
  

  distinct(coach, hc_name, college_clean)

cat("Coaching relationships identified:", nrow(coaching_relationships), "\n")
```

## Top Mentors

```{r mentors}
# Most prolific mentors (head coaches who produced assistants who became HCs)
prolific_mentors <- coaching_relationships |>
  filter(coach %in% d1_head_coaches) |>
  count(hc_name, sort = TRUE) |>
  rename(mentor = hc_name, proteges_who_became_hc = n)

cat("\nTop 20 mentors (head coaches whose assistants became D1 head coaches):\n")
print(prolific_mentors |> head(20), n = 20)
```

### 💡 STORY IDEAS - Section 3

**Based on the coaching trees data, brainstorm 2-3 story angles:**

1. The coaching tree data shows a small group of coaches --- led by Lynn Bria, Dawn Plitzuweit and Karen Aston --- consistently churning out future head coaches at a rate that dwarfs the rest of the sport.
2. Some programs aren’t just developing players, they’re developing coaches, with mentors like Kellie Harper, Michelle Clark-Heard and Brenda Frese running some of the most productive coaching pipelines in women’s basketball.
3. A coach’s most lasting legacy might not be wins, but the number of assistants they push into the head-coaching ranks.

**Questions to consider:**
- Who are the most prolific "mentor" coaches?
- Are there surprising names?
- Do certain coaches have "coaching trees" at multiple schools?
- Could you profile a specific coaching tree? Which one?

## DECISION POINT 5: Network Visualization Scope

**Question for you**: How many top mentors should we include in the visualization? Too many = cluttered, too few = missing important connections.

```{r network-visualization, fig.width=12, fig.height=10}
# STUDENT DECISION: Number of top mentors to visualize
N_MENTORS_TO_SHOW <- 10

# Create network of top mentors (for visualization clarity)
top_mentors <- prolific_mentors |>
  head(N_MENTORS_TO_SHOW) |>
  pull(mentor)

network_edges <- coaching_relationships |>
  filter(
    hc_name %in% top_mentors,
    coach %in% d1_head_coaches
  ) |>
  select(from = hc_name, to = coach, school = college_clean)

# Build network graph
coaching_network <- as_tbl_graph(network_edges, directed = TRUE)

# Plot
ggraph(coaching_network, layout = 'fr') +
  geom_edge_link(aes(alpha = 0.3), arrow = arrow(length = unit(2, 'mm'))) +
  geom_node_point(size = 5, color = "steelblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_void() +
  labs(
    title = "Coaching Trees: Top Mentors and Their Protégés",
    subtitle = paste("Top", N_MENTORS_TO_SHOW, "D1 head coaches who mentored future D1 head coaches")
  )
```

---

# Analysis 4: School-to-School Coaching Pipelines

```{r school-pipelines}
# Track coach movement between schools
coach_movements <- coaching |>
  filter(division == "I") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    next_school = lead(college_clean),
    next_position = lead(position_title_standardized),
    next_start = lead(start_year)
  ) |>
  filter(!is.na(next_school), college_clean != next_school) |>
  ungroup()

# Most common school-to-school transitions
cat("\nTop 30 school-to-school coaching pipelines:\n")
coach_movements |>
  count(college_clean, next_school, sort = TRUE) |>
  rename(from_school = college_clean, to_school = next_school, coaches = n) |>
  head(30) |>
  print(n = 30)
```

## Schools that Export to Head Coaching

```{r exports}
# Schools that "export" coaches to D1 head coaching jobs
coach_exports <- coach_movements |>
  filter(next_position == "Head Coach") |>
  count(college_clean, sort = TRUE) |>
  rename(school = college_clean, coaches_to_hc = n)

cat("\n\nTop 20 schools that export coaches to D1 head coaching positions:\n")
print(coach_exports |> head(20), n = 20)
```

### 💡 STORY IDEAS - Section 4

**Based on school-to-school movement patterns, brainstorm 2-3 story angles:**

1. The movement data shows how interconnected the sport really is, with surprising mid-majors, from Gardner-Webb to Northern Arizona, sending multiple coaches into bigger jobs and reshaping the coaching landscape.
2. The schools assistants come from, not just the ones they coach at, are shaping the hiring market, with certain institutions disproportionately supplying the next wave of head coaches.
3. Despite hundreds of D1 programs, most coach mobility routes funnel through a handful of schools, creating a tight circle where influence and opportunity keep recycling.

**Questions to consider:**
- Are there strong regional pipelines (e.g., schools from the same conference feeding each other)?
- Do assistant coaches at top programs tend to move to head coaching at mid-majors first?
- Are there unusual cross-country connections?
- Which schools are "feeders" vs. "destinations"?

---

# Analysis 5: Conference-Level Analysis

## DECISION POINT 6: Conference Focus

**Question for you**: Should we analyze all conferences equally, or focus on major/Power 5 conferences? Should we combine some smaller conferences?

```{r conference-analysis}
conference_production <- future_hc_positions |>
  filter(before_hc) |>         
  count(conference, position_title_standardized) |>
  arrange(desc(n))

cat("\nTop conferences producing future head coaches:\n")
conference_production |>
  group_by(conference) |>
  summarize(total = sum(n), .groups = "drop") |>
  arrange(desc(total)) |>
  head(15) |>
  print(n = 15)
```

## Conference Loyalty

```{r loyalty}

conference_loyalty <- coach_movements |>
  left_join(
    coaching |> 
      filter(division == "I") |>
      select(coach, college_clean, conference),
    by = c("next_school" = "college_clean", "coach" = "coach"),
    relationship = "many-to-many"
  ) |>
  filter(!is.na(conference.x), !is.na(conference.y)) |>
  mutate(same_conference = conference.x == conference.y) |>
  group_by(conference.x) |>
  summarize(
    total_moves = n(),
    same_conference_moves = sum(same_conference),
    loyalty_rate = same_conference_moves / total_moves * 100,
    .groups = "drop"
  ) |>
  arrange(desc(loyalty_rate))

cat("\n\nConference retention rates (coaches staying within conference):\n")
print(conference_loyalty |> head(15), n = 15)
```

### 💡 STORY IDEAS - Section 5

**Based on conference patterns, brainstorm 2-3 story angles:**

1. The ACC and SEC aren’t just winning games, they’re churning out the most future head coaches.
2. Coaches in the SWAC and Big Sky show the strongest loyalty in the country, staying within their leagues far more than their peers in the power conferences, where movement is constant and upward mobility is the norm.
3. The numbers make it clear there are two different coaching ecosystems: the high-powered conferences that export head coaches everywhere, and the smaller leagues where coaches tend to stay put and build long-term roots.

**Questions to consider:**
- Do certain conferences "develop" more coaches than others?
- Is there high loyalty within conferences, or do coaches move around?
- Are there conferences that import talent vs. export it?

---

# Analysis 6: Time to Head Coach Role

## DECISION POINT 7: Career Start Definition

**Question for you**: When calculating "time to head coach," should we count ALL positions (including high school, community college) or only college-level positions?

```{r time-to-hc}
time_to_hc <- coaching |>
  filter(coach %in% d1_head_coaches) |>
  filter(category == "College") |>  
  arrange(coach, start_year) |>
  group_by(coach) |>
  summarize(
    first_position_year = min(start_year, na.rm = TRUE),
    first_hc_year = min(start_year[position_title_standardized == "Head Coach"], na.rm = TRUE),
    years_to_hc = first_hc_year - first_position_year,
    .groups = "drop"
  ) |>
  filter(is.finite(years_to_hc), years_to_hc >= 0)

cat("\nTime to first head coach position:\n")
cat("Mean:", round(mean(time_to_hc$years_to_hc, na.rm = TRUE), 1), "years\n")
cat("Median:", median(time_to_hc$years_to_hc, na.rm = TRUE), "years\n")
cat("Min:", min(time_to_hc$years_to_hc, na.rm = TRUE), "years\n")
cat("Max:", max(time_to_hc$years_to_hc, na.rm = TRUE), "years\n")


ggplot(time_to_hc, aes(x = years_to_hc)) +
  geom_histogram(binwidth = 2, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = median(time_to_hc$years_to_hc, na.rm = TRUE), 
             linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Years to First Head Coaching Position",
    subtitle = "From first college coaching role to first head coach job",
    x = "Years",
    y = "Count of coaches"
  ) +
  theme_minimal()
```

### 💡 STORY IDEAS - Section 6

**Based on career timeline data, brainstorm 2-3 story angles:**

1. Becoming a head coach is a slow burn. Most coaches spend close to a decade climbing the ladder before they ever run their own program.
2. There’s no such thing as a “fast track.” A tiny handful of coaches get the job early, but the overwhelming majority grind through years of assistant roles first.
3. If you’re trying to map out the coaching pipeline, the pattern is clear. Time and reps matter. Experience, not shortcuts, is the currency that gets you a head-coaching shot.

**Questions to consider:**
- What's the typical timeline to becoming a D1 head coach?
- Who are the outliers (very fast or very slow)?

---

# Analysis 7: Elite Program "Finishing Schools"

## DECISION POINT 8: Defining "Elite Programs"

**Question for you**: Which schools should count as "elite" or "blue blood" programs? 
This is subjective - modify the list based on your knowledge or research.

```{r finishing-schools}

top_programs <- c("UConn", "Tennessee", "Stanford", "Notre Dame", "Baylor",
                  "South Carolina", "LSU", "Texas", "Duke", "UCLA",
                  "Louisville", "Maryland", "Ohio St.", "Iowa", "Indiana")


cat("Analyzing", length(top_programs), "programs\n")

# Assistant coaches from top programs who became head coaches
finishing_school_analysis <- future_hc_positions |>
  filter(
    before_hc,
    division == "I",
    college_clean %in% top_programs,
    position_title_standardized %in% c("Assistant Coach", "Associate Head Coach")
  ) |>
  group_by(college_clean) |>
  summarize(
    future_head_coaches = n_distinct(coach),
    .groups = "drop"
  ) |>
  arrange(desc(future_head_coaches))

cat("\n'Finishing schools' - top programs producing future head coaches:\n")
print(finishing_school_analysis, n = Inf)

# Average years spent at top program before becoming HC
years_at_top_programs <- future_hc_positions |>
  filter(
    before_hc,
    division == "I",
    college_clean %in% top_programs,
    position_title_standardized %in% c("Assistant Coach", "Associate Head Coach")
  ) |>
  mutate(years_at_school = coalesce(end_year, 2025) - start_year) |>
  group_by(college_clean) |>
  summarize(
    avg_years = mean(years_at_school, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(avg_years))

cat("\n\nAverage years spent at top programs (for future HCs):\n")
print(years_at_top_programs, n = Inf)
```

### 💡 STORY IDEAS - Section 7

**Based on elite program patterns, brainstorm 2-3 story angles:**

1. The pipeline out of “brand-name” programs isn’t uniform. Iowa and UConn assistants wait the longest before getting their own shot, while places like Maryland and Tennessee turn assistants into head coaches much faster.
2. If you want to become a head coach, picking the right powerhouse matters. The data suggests some bluebloods are true launchpads, and others require a much steeper climb.
3. Some programs groom future head coaches through long apprenticeships, while others churn out ready-made leaders in half the time.

**Questions to consider:**
- Do the biggest brand names produce the most head coaches?
- Are there "hidden gems" - programs that punch above their weight?
- Does time spent at elite programs matter (short stint vs. long apprenticeship)?
- Are there programs that hoard talent vs. those that develop and release it?

---

# Analysis 8: "Stepping Stone" Schools

```{r stepping-stones}
# Head coaches who left for another D1 head coaching job
stepping_stone_moves <- coaching |>
  filter(division == "I", position_title_standardized == "Head Coach") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    next_school = lead(college_clean),
    next_position = lead(position_title_standardized),
    moved_to_hc = next_position == "Head Coach" & !is.na(next_school)
  ) |>
  filter(moved_to_hc) |>
  ungroup()

cat("\nSchools most commonly serving as stepping stones (HC → HC moves):\n")
stepping_stone_moves |>
  count(college_clean, sort = TRUE) |>
  rename(school = college_clean, hc_departures = n) |>
  head(20) |>
  print(n = 20)

# Most common destination schools
cat("\n\nMost common destination schools for head coaches:\n")
stepping_stone_moves |>
  count(next_school, sort = TRUE) |>
  rename(destination = next_school, hc_arrivals = n) |>
  head(20) |>
  print(n = 20)
```

### 💡 STORY IDEAS - Section 8

**Based on stepping stone patterns, brainstorm 2-3 story angles:**

1. If you’re tracking the sport’s coaching carousel, these schools are where the cycle really starts. They’re the proving grounds --- the places ADs look when they want the next big hire.
2. Some schools have quietly become launchpads --- places where coaches land first, win just enough, and immediately get hired away by bigger jobs. The numbers show George Washington, Texas Tech and Wisconsin sit right at the center of that churn.
3. The data makes it clear there are “destination jobs” and “stepping-stone jobs,” and they aren’t always the ones you’d expect. A&M, Boston College and UCF show up as frequent landing spots for new head coaches rather than long-term homes.

**Questions to consider:**
- Which schools can't keep their successful head coaches?
- What makes a school a "destination" vs. a "stepping stone"?
- Are there schools that serve as consistent "first head coaching jobs"?
- Can you identify a "ladder" of schools (small to medium to major)?

---

# Analysis 9: Position Mobility Analysis

```{r position-mobility}
# Years spent in each position before moving up
position_duration <- coaching |>
  filter(division == "I") |>
  mutate(years_in_position = coalesce(end_year, 2025) - start_year) |>
  group_by(position_title_standardized) |>
  summarize(
    n_positions = n(),
    avg_duration = mean(years_in_position, na.rm = TRUE),
    median_duration = median(years_in_position, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(n_positions))

cat("\nAverage years in each position type:\n")
print(position_duration, n = Inf)

# Promotion rates by position
promotions <- coaching |>
  filter(division == "I") |>
  arrange(coach, start_year) |>
  group_by(coach) |>
  mutate(
    next_position = lead(position_title_standardized),
    promoted = case_when(
      position_title_standardized == "Staff" & 
        next_position %in% c("Assistant Coach", "Associate Head Coach", "Head Coach") ~ TRUE,
      position_title_standardized == "Assistant Coach" & 
        next_position %in% c("Associate Head Coach", "Head Coach") ~ TRUE,
      position_title_standardized == "Associate Head Coach" & 
        next_position == "Head Coach" ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  filter(!is.na(next_position)) |>
  group_by(position_title_standardized) |>
  summarize(
    total_moves = n(),
    promotions = sum(promoted),
    promotion_rate = promotions / total_moves * 100,
    .groups = "drop"
  ) |>
  arrange(desc(promotion_rate))

cat("\n\nPromotion rates by position:\n")
print(promotions, n = Inf)
```

### 💡 STORY IDEAS - Section 9

**Based on mobility and promotion patterns, brainstorm 2-3 story angles:**

1. If you want to climb in this profession, you need to be an assistant --- it’s the only on-ramp that consistently leads to a head job.
2. A handful of jobs are dead ends. Strength coaches, recruiting coordinators and interim titles show almost no upward movement, underscoring how narrow the true path to the big chair really is.
3. Staff roles quietly produce the most upward mobility, suggesting the pipeline starts even earlier than most people assume --- long before the “Assistant Coach” title ever appears.

**Questions to consider:**
- Which positions are "dead ends" vs. "springboards"?
- Do people spend more time in certain positions?
- Is there a "right" amount of time to spend before moving up?

---

# FINAL REFLECTION: Your Best Story Ideas

Review all your story ideas from each section. Which 3 would you pursue as full stories?

**Story #1:**

The assistant-to-head-coach track: Why nearly every D1 head coach takes the same long path.

**Why this story:**

The data makes the pipeline impossible to ignore --- the vast majority of future head coaches log years and years as assistants before even sniffing a top job. There’s a real story in how uniform the path is, how few shortcuts exist and how the grind has essentially become its own requirement.

**Story #2:**

Not All Bluebloods Are Equal

**Why this story:**

Not all bluebloods function the same. Some programs, like Maryland and Tennessee, turn assistants into head coaches quickly, while others demand long apprenticeships. That contrast is genuinely surprising and gives you a chance to show where coaches actually springboard from --- not just the brands fans assume create head coaches.

**Story #3:**
The sport’s true coaching factories are not the schools you’d expect.
**Why this story:**
A handful of mid-majors, places that never show up in Final Fours, are quietly producing more future head coaches than traditional powers. That flips the usual narrative on its head. It gives you a strong counterintuitive story: the biggest influence in shaping the coaching landscape isn’t coming from UConn or Stanford ...  it’s coming from places most fans couldn’t point to on a map.


# Additional Analysis Ideas to Explore

Now that you've worked through these analyses, what other questions could this data answer?

**Your ideas:**

1. Does winning actually matter for career mobility? Merge this with win-loss records or program success to see if coaches from winning staffs move up faster --- or if opportunity is more random than people think.
2. How often do coaches return to their alma mater — and does it help their career? We could track whether coaching at your alma mater accelerates or slows the path to a head-coaching job.
3. Which conferences hire internally the most — and which rely heavily on outsiders? We’ve already measured loyalty. We could take it further and look at which leagues recycle the same assistants vs. which bring in talent from everywhere.
