DEC 10 FINAL

The NFL didn’t overhaul the kickoff all at once. It nudged, experimented and finally committed to a full redesign.

In 2023, the league was still operating under the old world: long coverage run-ups, sky-high touchback rates and a play that had basically been legislated into irrelevance. 

Then came 2024, when the league borrowed the XFL model and rolled out the first version of the Dynamic Kickoff. It was a half-step — players aligned closer together, returns became more common and coaches suddenly had to think again — but it still felt like a trial run.

By 2025, the NFL locked in the rule permanently and sharpened it. 

Touchbacks to the 35 became standard if the ball landed in the end zone. Any kick that fell short of the landing zone was blown dead and spotted at the 40. Teams could declare an onside kick at any point, not just in the fourth quarter. And the spacing rules tightened, forcing coverage units to start flat-footed while returners gained a natural head start into open space.

The result is exactly what the league wanted, and what the data shows all over this report. 

Touchbacks collapsed, returns exploded and field position suddenly mattered again. The kickoff didn’t just come back to life; it became one of the most influential plays on the field almost overnight.

```{r}
library(nflfastR)
library(tidyverse)
library(lubridate)
library(scales)
```

```{r}
pbp <- load_pbp(2023:2025)
```

```{r}
kickoffs <- pbp |>
  filter(play_type == "kickoff") |>
  mutate(
    rule_phase = case_when(
      season < 2024 ~ "Old",
      season == 2024 ~ "Dynamic_Intro",
      season >= 2025 ~ "Dynamic_Refined"
    ),
    kickoff_type = case_when(
      touchback == 1 ~ "Touchback",
      kickoff_out_of_bounds == 1 ~ "Out of Bounds",
      kickoff_fair_catch == 1 ~ "Fair Catch",
      own_kickoff_recovery == 1 ~ "Onside Recovery",
      !is.na(return_yards) ~ "Return",
      TRUE ~ NA_character_
    ),
    result = case_when(
      touchback == 1 ~ "Touchback",
      !is.na(return_yards) ~ "Return",
      TRUE ~ "Other"
    )
  )
```

```{r}
drives <- pbp |>
  select(
    game_id,
    fixed_drive,
    fixed_drive_result,
    drive_start_yard_line,
    drive_end_yard_line,
    drive_ended_with_score
  ) |>
  distinct()
```

```{r}
kickoffs_drives <- kickoffs |>
  left_join(
    drives,
    by = c("game_id", "fixed_drive")
  )
```

```{r}
kickoffs_clean <- kickoffs_drives |>
  select(
    game_id,
    season,
    week,
    kicking_team = posteam,
    kickoff_type,
    kick_result = result,
    return_yards,
    kickoff_out_of_bounds,
    rule_phase,
    fixed_drive,
    drive_result = fixed_drive_result.y,
    drive_start_yard_line = drive_start_yard_line.y,
    drive_end_yard_line = drive_end_yard_line.y,
    drive_ended_with_score = drive_ended_with_score.y
  ) |>
  mutate(
    drive_start_yard_line = as.numeric(drive_start_yard_line),
    drive_end_yard_line   = as.numeric(drive_end_yard_line)
  )
```

```{r}
team_rates <- kickoffs_clean |>
  group_by(season, rule_phase, kicking_team) |>
  summarise(
    n_kickoffs = n(),
    pct_touchback = mean(kick_result == "Touchback", na.rm = TRUE),
    pct_return    = mean(kick_result == "Return",     na.rm = TRUE),
    pct_other     = mean(kick_result == "Other",      na.rm = TRUE),
    pct_return_inside25 = mean(
      kick_result == "Return" & drive_start_yard_line > 75,
      na.rm = TRUE
    ),
    avg_start_pos = mean(drive_start_yard_line, na.rm = TRUE),
    avg_end_pos   = mean(drive_end_yard_line,   na.rm = TRUE),
    scoring_rate_allowed = mean(drive_ended_with_score, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
league_rates <- kickoffs_clean |>
  group_by(season, rule_phase) |>
  summarise(
    total_kickoffs  = n(),
    pct_touchback   = mean(kick_result == "Touchback", na.rm = TRUE),
    pct_return      = mean(kick_result == "Return",     na.rm = TRUE),
    pct_other       = mean(kick_result == "Other",      na.rm = TRUE),
    pct_return_inside25 = mean(
      kick_result == "Return" & drive_start_yard_line > 75,
      na.rm = TRUE
    ),
    avg_start_pos = mean(drive_start_yard_line, na.rm = TRUE),
    avg_end_pos   = mean(drive_end_yard_line,   na.rm = TRUE),
    scoring_rate_allowed = mean(drive_ended_with_score, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
kickoff_type_rates <- kickoffs_clean |>
  group_by(season, rule_phase, kick_result) |>
  summarise(
    n = n(),
    avg_start_pos = mean(drive_start_yard_line, na.rm = TRUE),
    avg_end_pos   = mean(drive_end_yard_line,   na.rm = TRUE),
    scoring_rate_allowed = mean(drive_ended_with_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(season, rule_phase) |>
  mutate(
    pct_of_kickoffs = n / sum(n)
  ) |>
  ungroup()
```

```{r}
league_rates_long <- league_rates |>
  pivot_longer(
    cols = c(pct_touchback, pct_return, pct_other),
    names_to = "type",
    values_to = "rate"
  )

ggplot(league_rates_long, aes(
  x = season, y = rate, color = type, group = type
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Kickoff Returns Surge as \nTouchbacks Collapse Under the New Rule",
    subtitle = "Touchback vs Return vs Other, by Season",
    x = "Season",
    y = "Rate"
  ) +
  theme_minimal(base_size = 14)
```
The new kickoff basically erased the old decision tree. 

Before 2024, touchbacks dominated because coverage units had a 30-yard runway and returners rarely found space. The Dynamic Kickoff removes that runway entirely, forcing defenders to start flat-footed while the returner gets a head start and cleaner lanes. 

The result shows up immediately in the data: returns spike, touchbacks crash, and the “other” bucket — squibs, pooches, dead-ball plays --- shrinks because coaches no longer need those workarounds. The rule didn’t just tweak strategy. 

It rewired what a normal kickoff looks like, and by 2025 the league fully tilted toward live returns as the default outcome.

```{r}
kickoff_scoring <- kickoffs_clean |>
  group_by(season, kickoff_type) |>
  summarise(
    scoring_rate = mean(drive_ended_with_score, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(kickoff_scoring, aes(
  x = factor(season),
  y = scoring_rate,
  fill = kickoff_type
)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Touchbacks aren’t the safe play anymore, \nneither are touchbacks",
    subtitle = "Returns in 2025 produce more points than touchbacks in 2023",
    x = "Season",
    y = "Scoring Rate",
    fill = "Kickoff Type"
  ) +
  theme_minimal(base_size = 14)
```

That onside kick scoring rate looks weird. Let's check that. 

```{r}
kickoffs_clean |>
  count(season, kickoff_type)
```

```{r}
kickoffs_clean |>
  filter(kickoff_type == "Onside Recovery") |>
  select(season, game_id, drive_ended_with_score, drive_start_yard_line, drive_end_yard_line)
```

```{r}
kickoffs_clean |>
  filter(kickoff_type == "Onside Recovery") |>
  group_by(season) |>
  summarise(
    attempts = n(),
    scoring_rate = mean(drive_ended_with_score, na.rm = TRUE)
  )
```

```{r}
onside_plays <- pbp |>
  filter(play_type == "kickoff",
         own_kickoff_recovery == 1) |>
  select(season, game_id, play_id, desc,
         drive, fixed_drive, drive_start_yard_line, drive_end_yard_line,
         drive_ended_with_score)

onside_plays
```

```{r}
pbp |>
  filter(game_id == "2024_12_MIN_CHI") |>
  arrange(play_id) |>
  mutate(next_play = lead(desc)) |>
  filter(play_id == 4189)
```

It checks out! Looks like the momentum of recovering it resulted in a score those seasons. 

Now back to the original graphic without that distraction!


```{r}
kickoff_scoring_no_onside <- kickoffs_clean |>
  filter(kickoff_type != "Onside Recovery") |>
  group_by(season, kickoff_type) |>
  summarise(
    scoring_rate = mean(drive_ended_with_score, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(kickoff_scoring_no_onside, aes(
  x = factor(season),
  y = scoring_rate,
  fill = kickoff_type
)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Touchbacks aren’t the safe play anymore, \nneither are touchbacks",
    subtitle = "Shows true scoring impact without high-variance onside recoveries",
    x = "Season",
    y = "Scoring Rate",
    fill = "Kickoff Type"
  ) +
  theme_minimal(base_size = 14)
```

Under the old rules, a touchback was the safe, boring answer --- flip the field, get the ball at the 25, avoid disaster. 

The Dynamic Kickoff flipped that logic on its head. 

Because returners now catch the ball closer to a spread-out coverage unit, they routinely push past the old “safe” touchback yardage. That means giving a team a free 25-yard start is no longer the conservative option; it’s actually worse field position than what most returns produce. 

That’s why return plays jump ahead in scoring rate and touchbacks slide backward. The data shows what coaches already learned in real time: the safest call used to be taking the knee. 

Now, it’s often the more dangerous one.

```{r}
pbp_old <- load_pbp(2015:2022)
```

```{r}
kickoffs_old <- pbp_old |>
  filter(play_type == "kickoff") |>
  mutate(
    rule_phase = "Pre-Dynamic",
    kickoff_type = case_when(
      touchback == 1 ~ "Touchback",
      kickoff_out_of_bounds == 1 ~ "Out of Bounds",
      kickoff_fair_catch == 1 ~ "Fair Catch",
      own_kickoff_recovery == 1 ~ "Onside Recovery",
      !is.na(return_yards) ~ "Return",
      TRUE ~ NA_character_
    ),
    result = case_when(
      touchback == 1 ~ "Touchback",
      !is.na(return_yards) ~ "Return",
      TRUE ~ "Other"
    )
  )
```

```{r}
drives_old <- pbp_old |>
  select(
    game_id,
    fixed_drive,
    fixed_drive_result,
    drive_start_yard_line,
    drive_end_yard_line,
    drive_ended_with_score
  ) |>
  distinct()
```

```{r}
kickoffs_drives_old <- kickoffs_old |>
  left_join(
    drives_old,
    by = c("game_id", "fixed_drive")
  )
```

```{r}
kickoffs_clean_old <- kickoffs_drives_old |>
  select(
    game_id,
    season,
    week,
    kicking_team = posteam,
    kickoff_type,
    kick_result = result,
    return_yards,
    rule_phase,
    fixed_drive,
    drive_result = fixed_drive_result.y,
    drive_start_yard_line = drive_start_yard_line.y,
    drive_end_yard_line = drive_end_yard_line.y,
    drive_ended_with_score = drive_ended_with_score.y
  ) |>
  mutate(
    drive_start_yard_line = as.numeric(drive_start_yard_line),
    drive_end_yard_line   = as.numeric(drive_end_yard_line)
  )
```

```{r}
league_scoring <- bind_rows(
  kickoffs_clean_old,
  kickoffs_clean 
) |>
  group_by(season) |>
  summarise(
    scoring_rate = mean(drive_ended_with_score, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
ggplot(league_scoring, aes(
  x = season,
  y = scoring_rate
)) +
  geom_line(color = "steelblue", linewidth = 1.3) +
  geom_point(color = "steelblue", size = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks = 2015:2025) +
  labs(
    title = "Scoring rate has risen with kickoff changes",
    subtitle = "Percentage of drives that ended in points following a kickoff",
    x = "Season",
    y = "Scoring Rate"
  ) +
  theme_minimal(base_size = 14)
```

The scoring bump tracks right alongside the league’s kickoff experiments. 

As touchbacks fell and returns became the default outcome, offenses started inheriting shorter fields more often. A return that crosses the 30 or 35 used to be a bonus—now it’s routine. 

Even when returners don’t break one, the new spacing forces more open-field plays, missed tackles, and chunk gains before the defense can get organized. 

All of that pushes expected points higher. 

The Dynamic Kickoff was designed to create more live action; the rise in scoring rate shows the downstream impact: more returns, better starting field position, and more drives ending in points.

```{r}
kickoffs_clean |>
  count(season, kickoff_type) |>
  group_by(season) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(
    x = kickoff_type,
    y = pct,
    fill = kickoff_type
  )) +
  geom_col() +
  facet_wrap(~ season) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Returns are now the dominant outcome",
    x = "Kickoff Type",
    y = "Percentage"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The shift in outcomes is the clearest fingerprint of the Dynamic Kickoff. 

Under the new system, almost every ball that lands in the field of play has to be returned — fair catches are essentially eliminated, and touchbacks only happen on balls that reach or cross the end zone. That forces action. 

Before the rule, coverage units had time and space to get downfield, so returners often waved it off and settled for guaranteed field position. Now the returner is catching the ball with ten defenders already stacked near the 40, turning every kickoff into a live play by design. 

The spike in returns --- and the collapse of touchbacks --- is exactly what the league intended: more football, fewer dead plays, and a completely different risk-reward profile for special teams.

```{r}
oob_totals <- kickoffs_clean |>
  filter(kickoff_out_of_bounds == 1) |>
  group_by(season) |>
  summarise(
    total_oob_kickoffs = n(),
    .groups = "drop"
  )
```

```{r}
ggplot(oob_totals, aes(x = factor(season), y = total_oob_kickoffs)) +
  geom_col(fill = "red") +
  labs(
    title = "Out-of-Bounds Kickoffs Spike Under the NFL’s New Rule",
    x = "Season",
    y = "Total OOB Kickoffs"
  ) +
  theme_minimal(base_size = 14)
```

Out-of-bounds kicks climbed as teams experimented with the new spacing of the Dynamic Kickoff. With coverage units starting so close to the returner, more coaches tried to steal hidden yards by pinning teams deep or testing directional kicks toward the sideline. 

The margin for error shrinks on those shots, and the new rule punishes misses harshly — one yard too wide turns an aggressive kickoff into a free spot near the 40. The jump in OOB kicks reflects that learning curve: teams pushing for an edge, and the rule pushing back.

```{r}
kickoffs_clean |>
  filter(kickoff_type %in% c("Touchback", "Out of Bounds", "Return")) |>
  group_by(season, kickoff_type) |>
  summarise(
    min_yardline = min(true_start, na.rm = TRUE),
    avg_yardline = mean(true_start, na.rm = TRUE),
    median_yardline = median(true_start, na.rm = TRUE),
    max_yardline = max(true_start, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(season, kickoff_type)
```
```{r}
kickoffs_clean |>
  filter(kickoff_type == "Touchback") |>
  count(season, true_start)
```

```{r}
kickoffs_clean |>
  filter(kickoff_type == "Out of Bounds") |>
  count(true_start)
```

```{r}
faircatch_oob <- kickoffs_clean |>
  count(season, kickoff_type) |>
  filter(kickoff_type %in% c("Fair Catch", "Out of Bounds"))

ggplot(faircatch_oob, aes(
  x = factor(season),
  y = n,
  fill = kickoff_type
)) +
  geom_col(position = "dodge") +
  labs(
    title = "Fair Catches Decline While Out-of-Bounds Kicks Rise",
    subtitle = "The new rule removes the fair-catch option entirely, \npushing teams toward risk-free out-of-bounds attempts instead.",
    x = "Season",
    y = "Number of Kickoffs",
    fill = "Kickoff Type"
  ) +
  theme_minimal(base_size = 14)
```
Fair catches all but vanished once the NFL moved to the Dynamic Kickoff. 

The old rule gave returners an easy escape hatch — signal, catch it, and take the free field position. The new system removes that option entirely. Any ball that lands in the “landing zone” has to be returned, and anything caught in the end zone becomes a touchback by rule. 

There’s no discretion built into the play anymore. So the drop in fair catches isn’t a coaching trend or a shift in philosophy; it’s simply the byproduct of a rule designed to eliminate them and force the ball back into play.

