```{r}
library(tidyverse)
```


```{r}
logs <- read_csv("https://dwillis.github.io/sports-data-files/footballlogs1124.csv") |>
   mutate(PointDifferential = TeamScore - OpponentScore)

logs |> select(Team, Opponent, TeamScore, OpponentScore, PointDifferential) |>
  head()
```

```{r}
model <- lm(PointDifferential ~ PenaltyYds, data = logs)

summary(model)

```

***

The regression shows a coefficient of 0.015 for penalty yards, with a p-value of 0.0067. Since this is below 0.05, the relationship is statistically significant, meaning it is unlikely to be the result of random chance. The R-squared value is only 0.0003, meaning penalty yards explain less than 1% of the variation in score differential. While the model detects a statistically real relationship, it has no predictive power—penalty yards alone are not a useful way to explain or predict game outcomes.

***

```{r}
multiple_regression_model <- lm(PointDifferential ~ Penalties + PenaltyYds + Fumbles + Interceptions + TotalTurnovers, data = logs)

summary(multiple_regression_model)
```
***
The multiple regression is significant, with penalties, fumbles and interceptions explaining about 14% of score differential. TotalTurnovers dropped due to multicollinearity, showing overlap among predictors. Mistakes matter but most variance remains unexplained.
***

```{r}
close_games <- logs |>
  filter(PointDifferential <= 7 & PointDifferential >= -7)
```

```{r}
multiple_regression_close_games_model <- lm(PointDifferential ~ Penalties + PenaltyYds + Fumbles + Interceptions, data = close_games)

summary(multiple_regression_close_games_model)
```

***
In close games, the multiple regression is stronger than the simple model, but still only explains about 5% of score differential. Mistakes matter, yet not enough for prediction. It is better used as context showing penalties and turnovers can tilt tight games.
***

***
Penalties and turnovers correlate with point differential but explain little on their own, making them weak predictors. Still, they’re useful context. It is a stat to support the idea that mistakes swing games, but would be best used as a sidebar or nugget within a larger story.
***

