```{r}
library(tidyverse)
```

```{r}
soccer <- read_csv("https://dwillis.github.io/sports-data-files/ncaa_womens_soccer_totals_2024.csv")
```

```{r}
soccer |>
  mutate(SoGPct = (SoG/ShAtt)*100
) |>
  arrange(desc(SoGPct))
```

```{r}
soccer |>
  mutate(
    SoGPct = (SoG/ShAtt) * 100,
    GoalsPerGame = goals / games_played
  ) |>
  arrange(desc(GoalsPerGame)) |>
  select(school_name, conference_name, GoalsPerGame)
```

```{r}
soccer |>
  mutate(
    GoalsPerGame = goals / games_played
  ) |>
  arrange(desc(goalie_sv_pct)) |>
  select(school_name, conference_name, GoalsPerGame, goalie_sv_pct)
```


Four of the top 10 teams in the country in shot-on-goal percentage come from the Southwestern Athletic Conference. For a league not exactly known for its soccer pedigree, that’s a surprising sight at the top of the leaderboard.

But the numbers don’t tell a simple story. None of those teams crack the top 25 nationally in goals per game, which hints at a different dynamic: maybe defenses in the SWAC aren’t closing down shooters as tightly, giving attackers cleaner looks, while goalkeepers are doing just enough to keep scores down. It paints a picture of bad defending paired with solid keeping — an unusual balance that helps explain the efficiency gap.

The next step is figuring out how far this trend extends. Are defenses across the SWAC weaker than other mid-major leagues? And what happens when players transfer out — do they keep producing quality shots at the same rate in tougher conferences, or does the efficiency fade once the defensive level rises? Those are the questions that could turn a quirky stat into a deeper story about how talent in the SWAC translates beyond. 

