---
title: "Background Research"
author: "Michael Howes"
format: html
---

***This is the question I chose: Does higher-than-average home attendance improve the home team’s performance in college football games, such as Ohio State/Michigan coming to play at SECU Stadium compared to a Towson. Does that help, or actually hurt the Terps' performance.  ****


```{r}
library(tidyverse)
```

```{r}
data <- read_csv("../data/attendance spreadsheet.csv")
```
```{r}
head(data)
```


```{r}
summary(data)
```

```{r}
data |>
  group_by(Result) |>
  summarise(avg_md_turnovers = mean(MdTurnovers),
            avg_opp_turnovers = mean(OppTurnovers),
            games = n())
```

```{r}
data <- data |>
  rename(Attendance = Attendence)
```


```{r}
data |>
  group_by(Result) |>
  summarise(avg_attendance = mean(Attendance))
```

```{r}
data |> 
  filter(Attendance > 38000) |> 
  summarise(avg_md_points = mean(MdPoints),
            avg_opp_points = mean(OppPoints),
            wins = sum(Result == "W"),
            losses = sum(Result == "L"),
            games = n())
```

```{r}
data |> 
  filter(Attendance <= 38000) |> 
  summarise(avg_md_points = mean(MdPoints),
            avg_opp_points = mean(OppPoints),
            wins = sum(Result == "W"),
            losses = sum(Result == "L"),
            games = n())
```


***I found that Maryland averaged about 35 points and went 11–3 in games with below-average attendance, compared to just 24 points and a 2–5 record in high-attendance games. The results suggest that larger crowds at SECU Stadium don’t translate into better performance for Maryland — instead, those games typically feature ranked Big Ten opponents who draw more fans but also hand Maryland losses.***